---
title: "Data Cleaning and Demostration"
date: "2025-03-17"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(reshape2)
library(gridExtra)
```

# Importing the Dataset and Column Descriptions  

First, we import the dataset and provide column descriptions for better understanding.

```{r}
Violence_raw <- read_csv("Violence_Reduction_-_Victim_Demographics_-_Aggregated.csv", show_col_types = FALSE)
Violence_raw$Year <- substr(Violence_raw$TIME_PERIOD, 1, 4)
```

**TIME_PERIOD**  
The time period aggregated.

**TIME_PERIOD_START**  
All victimizations have been aggregated to the quarter level. This time stamp reflects the start of the quarter.

**TIME_PERIOD_END**  
All victimizations have been aggregated to the quarter level. This time stamp reflects the end of the quarter.

**PRIMARY_TYPE**  
Text description of the IUCR Code that describes the major crime category for the corresponding victimization it falls into. This column represents only the crime that this specific victim within the incident experienced.

**AGE**  
The age of the victims grouped by decade.

**SEX**  
The sex of the victims. Victimization demographic data shown here are captured by CPD and limited to information included in the police report, which sometimes is not supported by individual self-reporting.

**RACE**  
The race of the victims. Victimization demographic data shown here are captured by CPD and limited to information included in the police report, which may not often be supported by individual self-reporting.

**JUVENILE_I**  
Whether the victims were juveniles (under age 18). A blank value means Unknown.

**DOMESTIC_I**  
Indicates whether the aggregated number of victimizations were domestic-related as determined by the initial reporting officer or the investigating detective.

**GUNSHOT_INJURY_I**  
Indicator field describing whether or not a victim was injured by gunfire. Shooting data is not available before 2010 so all non-homicide victimizations prior to 2010 will be recorded as "UNKNOWN".

**NUMBER_OF_VICTIMS**  
The number of victims matching the unique combination of all other columns.





# Data Cleaning: Handling Missing Values  

Next, we perform data cleaning. The first step is to remove potential missing values. The columns that may contain missing values are as follows:  

**"AGE", "SEX", "RACE", "JUVENILE_I", "DOMESTIC_I", "GUNSHOT_INJURY_I", "NUMBER_OF_VICTIMS", "PRIMARY_TYPE"**  

We will now count the number of missing values in each of these columns.
```{r}
Violence_na <- Violence_raw[apply(Violence_raw[, c("AGE", "SEX", "RACE", "JUVENILE_I", 
                                                   "DOMESTIC_I", "GUNSHOT_INJURY_I", 
                                                   "NUMBER_OF_VICTIMS", "PRIMARY_TYPE")], 
                                  1, function(row) any(row == "UNKNOWN")), ]

Violence_remove_na <- Violence_raw[!apply(Violence_raw[, c("AGE", "SEX", "RACE", "JUVENILE_I", 
                                                            "DOMESTIC_I", "GUNSHOT_INJURY_I", 
                                                            "NUMBER_OF_VICTIMS", "PRIMARY_TYPE")], 
                                           1, function(row) any(row == "UNKNOWN")), ]


Violence_remove_na <- Violence_remove_na %>%
  mutate(Year = as.numeric(Year))

unknown_counts <- colSums(Violence_raw[, c("AGE", "SEX", "RACE", "JUVENILE_I", 
                                           "DOMESTIC_I", "GUNSHOT_INJURY_I", 
                                           "NUMBER_OF_VICTIMS", "PRIMARY_TYPE")] == "UNKNOWN")


unknown_df <- data.frame(Category = names(unknown_counts), Count = unknown_counts)

ggplot(unknown_df, aes(x = Category, y = Count, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(title = "Count of UNKNOWN Values in Each Category",
       x = "Category",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
We found that **GUNSHOT_INJURY_I** has an extremely high number of missing values. However, according to the dataset definition:  

*"Shooting data is not available before 2010, so all non-homicide victimizations prior to 2010 will be recorded as 'UNKNOWN'."*  

It seems that **UNKNOWN** in records before 2010 should not be considered as missing values. We will remove these "UNKNOWN" values from **GUNSHOT_INJURY_I** and then re-evaluate the missing values.  

At the same time, we found that missing values exist only in **"AGE"**, **"SEX"**, **"RACE"**, and **"GUNSHOT_INJURY_I"**.  

In the following sections, we will focus our analysis on these four columns.  

Next, we will visualize the results by generating two plots:  
1. The distribution of missing values per row.  
2. The distribution of missing values per column.  

```{r}
Violence_na <- Violence_na %>%
  mutate(Year = as.numeric(Year),
         GUNSHOT_INJURY_I = ifelse(Year <= 2010 & GUNSHOT_INJURY_I == "UNKNOWN", 
                                   "Before2010", 
                                   GUNSHOT_INJURY_I))

Violence_na$missing_count <- rowSums(Violence_na == "UNKNOWN")

missing_summary <- table(Violence_na$missing_count)

variable_missing_counts <- colSums(Violence_na == "UNKNOWN")[c("AGE", "SEX", "RACE", "GUNSHOT_INJURY_I")]

missing_summary_df <- data.frame(Missing_Count = as.numeric(names(missing_summary)), 
                                 Count = as.numeric(missing_summary))

variable_missing_df <- data.frame(Variable = names(variable_missing_counts), 
                                  Count = as.numeric(variable_missing_counts))

p1 <- ggplot(missing_summary_df, aes(x = factor(Missing_Count), y = Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Distribution of UNKNOWN Values per Row", 
       x = "Number of UNKNOWN Values", 
       y = "Row Count") +
  theme_minimal()

p2 <- ggplot(variable_missing_df, aes(x = Variable, y = Count)) +
  geom_bar(stat = "identity", fill = "coral") +
  labs(title = "Number of UNKNOWN Values per Variable", 
       x = "Variable", 
       y = "Count of UNKNOWN Values") +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)
```

We found that the missing values in **"GUNSHOT_INJURY_I"** have been reduced to zero. This suggests that all missing values in **"GUNSHOT_INJURY_I"** were caused by the dataset's counting rules.  

Among **"AGE"**, **"SEX"**, and **"RACE"**, having a single missing value might be acceptable. Before analyzing rows with two missing values, we refer to the Chicago Police Department's definitions of **"UNKNOWN"** in **"SEX"** and **"RACE"** as follows:  
(https://www.cpdwiki.org/wiki/quick_reference/race_codes)

*Race Definitions*  

- **BLK**: A person having origins in any black racial groups of Africa.  
- **WHI**: A person having origins in any of the peoples of Europe, North Africa, or the Middle East.  
- **WBH**: A person having origins in any black racial groups of Africa **and** is of Mexican, Puerto Rican, Cuban, Central or South American, or other Spanish culture or origin.  
- **WWH**: A person having origins in any of the peoples of Europe, North Africa, or the Middle East **and** is of Mexican, Puerto Rican, Cuban, Central or South American, or other Spanish culture or origin.  
- **I**: A person having origins in any of the original peoples of North America and who maintains cultural identification through tribal affiliation or community recognition.  
- **API**: A person having origins in any of the original peoples of the Far East, Southeast Asia, the Indian subcontinent, or the Pacific Islands; this area includes, for example, China, India, Japan, Korea, the Philippine Islands, Samoa, etc.  
- **Unknown**: A person whose racial and/or ethnic origin is unknown or one who refuses to disclose their racial and/or ethnic origin.  


*Gender Definitions*  

- **Male (M)**  
- **Female (F)**  
- **Non-binary (N)**  
- **Unknown or declined to answer (X)**  




We are curious about how rows with two missing values are distributed, so we visualize the missing value patterns using a heatmap.



```{r}
Violence_na$missing_count <- rowSums(Violence_na == "UNKNOWN")
df_missing_2 <- Violence_na[Violence_na$missing_count == 2, ]

vars <- c("AGE", "SEX", "RACE", "GUNSHOT_INJURY_I")
missing_matrix <- matrix(0, nrow = 4, ncol = 4, dimnames = list(vars, vars))

for (i in 1:nrow(df_missing_2)) {
  missing_vars <- names(df_missing_2[i, ])[df_missing_2[i, ] == "UNKNOWN"]
  if (length(missing_vars) == 2) {
    missing_matrix[missing_vars[1], missing_vars[2]] <- missing_matrix[missing_vars[1], missing_vars[2]] + 1
    missing_matrix[missing_vars[2], missing_vars[1]] <- missing_matrix[missing_vars[2], missing_vars[1]] + 1
  }
}

heatmap_data <- as.data.frame(as.table(missing_matrix))
colnames(heatmap_data) <- c("Variable1", "Variable2", "Count")

ggplot(heatmap_data, aes(x = Variable1, y = Variable2, fill = Count)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "red") +
  labs(title = "Heatmap of Missing Value Pairs", 
       x = "Variable", 
       y = "Variable", 
       fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
All cases with two missing values result from suspects refusing to provide information. Since this data accounts for only around a thousand records, we consider it of limited value and will exclude these cases from our analysis.


```{r}
Violence_na <- Violence_na %>%
  mutate(GUNSHOT_INJURY_I = ifelse(GUNSHOT_INJURY_I == "Before2010", "UNKOWN", GUNSHOT_INJURY_I))
Violence_na$missing_count <- rowSums(Violence_na == "UNKNOWN")
Violence_filtered <- Violence_na %>% filter(missing_count < 2) %>% select(-missing_count)
Violence_clean_missing <- bind_rows(Violence_remove_na, Violence_filtered)
```

# Comparing Chicago Crime Data by Race with U.S. Census Data  

Next, we will compare the racial distribution in Chicago crime data with the racial demographics from the U.S. Census for Chicago.  

The crime dataset spans from **1991 to 2025**, while census data is available for **1990, 2000, 2010, and 2020**. To align these datasets, we will group the crime data using **10-year** and **5-year intervals** for analysis.

```{r}
Violence_clean_missing$Year <- as.numeric(Violence_clean_missing$Year)
year_ranges_5 <- split(Violence_clean_missing, cut(Violence_clean_missing$Year, 
                        breaks = seq(1991, 2026, by = 5), right = FALSE, labels = FALSE))
year_ranges_10 <- split(Violence_clean_missing, cut(Violence_clean_missing$Year, 
                         breaks = c(seq(1991, 2021, by = 10), 2026), right = FALSE, labels = FALSE))
for (i in seq_along(year_ranges_5)) {
  assign(paste0("Violence_5yr_", i), year_ranges_5[[i]])
}
for (i in seq_along(year_ranges_10)) {
  assign(paste0("Violence_10yr_", i), year_ranges_10[[i]])
}
```
Here is the racial distribution of crime records in Chicago across all available years. 

If nationwide racial crime distribution data is available, we can add it for comparison.

```{r,echo=FALSE,message=FALSE}
temp_1 <- Violence_clean_missing
temp_1 <- temp_1 %>% filter(RACE != "UNKNOWN")
race_counts <- temp_1 %>%
  group_by(RACE) %>%
  summarise(Count = n()) %>%
  mutate(Proportion = Count / sum(Count) * 100)
ggplot(race_counts, aes(x = RACE, y = Proportion, fill = RACE)) +
  geom_bar(stat = "identity") +
  labs(title = "Proportion of Each Race", 
       x = "Race", 
       y = "Percentage of Total") +
  theme_minimal() +
  theme(legend.position = "none")
rm(temp_1)
gc()  
```
**Here is data from (https://www.census.gov)**

```{r}
chicago_demographics <- data.frame(
  Year = c(1990, 2000, 2010, 2020),
  WHI= c(37.9, 31.3, 31.7, 31.4),
  BLK = c(39.1, 36.8, 32.3, 28.7),
  HIS = c(19.6, 26.0, 28.9, 29.8),
  API = c(3.7, 4.3, 5.4, 6.9),
  I = c(0.2,0.15,0.15,0.12)
)
chicago_demographics
```

Below is a comparison of racial crime record proportions and U.S. Census results, with a separate statistical chart for each decade.
```{r}
plot_comparison <- function(violence_df, year, title) {
  filtered_df <- violence_df %>% filter(!(RACE %in% c("UNKNOWN", "S")))
  
  # Adjust HIS category by combining WBH and WWH into HIS
  filtered_df <- filtered_df %>%
    mutate(RACE = ifelse(RACE %in% c("WBH", "WWH"), "HIS", RACE))
  
  race_counts <- filtered_df %>%
    group_by(RACE) %>%
    summarise(Count = n()) %>%
    mutate(Proportion = Count / sum(Count) * 100)
  
  chicago_data <- chicago_demographics[chicago_demographics$Year == year, ]
  chicago_proportions <- data.frame(
    RACE = c("WHI", "BLK", "HIS", "API", "I"),
    Proportion = as.numeric(chicago_data[2:6])
  )
  
  race_counts$Source <- "Violence Data"
  chicago_proportions$Source <- "Census Data"
  combined_data <- bind_rows(race_counts, chicago_proportions)
  
  ggplot(combined_data, aes(x = RACE, y = Proportion, fill = Source)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(title = title, x = "Race", y = "Proportion (%)") +
    theme_minimal()
}

p1 <- plot_comparison(Violence_10yr_1, 1990, "Comparison: 1990")
p2 <- plot_comparison(Violence_10yr_2, 2000, "Comparison: 2000")
p3 <- plot_comparison(Violence_10yr_3, 2010, "Comparison: 2010")
p4 <- plot_comparison(Violence_10yr_4, 2020, "Comparison: 2020")

grid.arrange(p1, p2, p3, p4, ncol = 2, nrow = 2)
```

```{r}
Violence_clean_missing
```

```{r}
# Import library
library(dplyr)
library(ggplot2)
library(broom)
library(ggeffects)  #

# Binary variable
Violence_model_data <- Violence_clean_missing %>%
  filter(PRIMARY_TYPE != "UNKNOWN", SEX != "UNKNOWN", RACE != "UNKNOWN") %>%
  mutate(
    IS_SEXUAL_ASSAULT = ifelse(PRIMARY_TYPE == "CRIMINAL SEXUAL ASSAULT", 1, 0),
    SEX = factor(SEX),
    RACE = factor(RACE)
  )

# Logistic regression model 
model <- glm(IS_SEXUAL_ASSAULT ~ SEX * RACE, 
             data = Violence_model_data, 
             family = binomial)

# model summary
summary(model)  
tidy(model, exponentiate = TRUE, conf.int = TRUE)  

# probability predict
library(ggeffects)
pred <- ggpredict(model, terms = c("SEX", "RACE"))

# visuaization
plot(pred) +
  labs(title = "Predicted Probability of Being a Sexual Assault Victim",
       x = "Sex",
       y = "Predicted Probability") +
  theme_minimal()

```

